Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0

### 2019-06-10 2019_04_22_Reviewer_Comments.Rmd
re-submission commit!

Note to all reviewers:
Reviewer 2 asked why we had not included the Ratnapriya et al. large AMD retina RNA-seq study. The reason was because the data was made available after our last data pull. While addressing the reviewer criticisms we decided to pull new GTEx samples along with all eye tissue datasets as of May 2019. This will change all sample counts in the document. We **have** corrected the incorrect code that generated the wrong sample numbers.

@@

Zhang et al. (https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-017-4002-1) report a pearson cor R^2 with TPM of 0.89 and 0.93 between alignment free (salmon/kallisto) methods compared to alignment-based methods (STAR/REM) [@zhangEvaluationComparisonComputational2017]. As GTEx used an alignment-based approach (and a different reference genome - we use GRCh38, GTEx uses GRCh37) and we length scale our TPM values (GTEx does note) we feel that this is a reasonable result.

We have added a section to our methods and results section discussing this briefly (lines 126-129, 227-233).

@@

We have added a "limitations" section to the resulst section of the manuscript:

>>>>

Note to all reviewers:
Reviewer 2 asked why we had not included the Ratnapriya et al. large AMD retina RNA-seq study [@ratnapriyaRetinalTranscriptomeEQTL2019]. The reason was because the data was made available after our last data pull. While addressing the reviewer criticisms we decided to pull new GTEx samples along with all eye tissue datasets as of May 2019. This will change all sample counts in the document. We **have** corrected the incorrect code that generated the wrong sample numbers.

@@

To ensure there are no substantial differences in quantification of gene TPM values, we calculated the $R^2$ between GTEX and EiaD generated TPM values for our shared GTEx samples (see methods); we computed an $R^2$ of 0.89. Zhang et al. report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a $R^2$ ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017]. As Zhang et al. compared quantification methods with identical gene references (we use Gencode GRCh38 gene models and GTEx uses hg19) and did not scale TPM score differently, our result falls in line with expectations.

We have added a section to our methods and results section discussing this briefly (lines 126-129, 227-232).

@@

We have added a "limitations" section to the results section of the manuscript:

### 2019-06-10 manuscript.Rmd
re-submission commit!

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To develop an easily accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 1375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples. The 916 eye and 1375 GTEx samples were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD. RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date by database version, gene/transcript, 19 eye tissues, and 54 body tissues. As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retina are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families like protocadherin and HOXB family members. CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

To ensure there are no substantial differences in quantification of gene TPM values, we determined the $R^2$ between GTEX and EiaD generated TPM values for our shared GTEx samples (see methods); we computed an $R^2$ of 0.89. “ Zhang et al. report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a $R^2$ ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017]. As Zhang et al. compared quantification methods with identical gene references (we use Gencode GRCh38 gene models and GTEx uses hg19) and did not scale TPM score differently, our result falls in line with expectations.

>>>>

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To develop an accessible and reliable RNA-seq transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 1375 RNA-seq samples across 54 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non-perturbed, human eye-related tissue RNA-seq samples. The 916 eye and 1375 GTEx samples were sent into a Snakemake-based reproducible pipeline we wrote to quantify all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 882 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. Furthermore, we rewrote the web application eyeIntegration (https://eyeIntegration.nei.nih.gov) to display EiaD. RESULTS: The new eyeIntegration portal provides quick visualization of human eye-related transcriptomes published to date by database version, gene/transcript, 19 eye tissues, and 54 body tissues. As a test of the value of this unified pan-eye dataset, we showed that fetal and organoid retina are highly similar at a pan-transcriptome level but display distinct differences in certain pathways and gene families like protocadherin and HOXB family members. CONCLUSION: The eyeIntegration v1.0 web app serves the pan-human eye and body transcriptome dataset, EiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

@@

To ensure there are no substantial differences in quantification of gene TPM values, we calculated the $R^2$ between GTEX and EiaD generated TPM values for our shared GTEx samples (see methods); we computed an $R^2$ of 0.89. Zhang et al. report that RNA-seq quantifications done between alignment-free methods (used in EiaD) and alignment-based methods (used by GTEx) get a $R^2$ ranging from 0.89 to 0.93 [@zhangEvaluationComparisonComputational2017]. As Zhang et al. compared quantification methods with identical gene references (we use Gencode GRCh38 gene models and GTEx uses hg19) and did not scale TPM score differently, our result falls in line with expectations.
